### 2022-06-19 chapter1/chapter.tex
Ch7 2nd draft

A stochastic process is a sequence, \(\{X(t)\}_{t\in\mathcal T}\), of random variables (or random vectors) indexed by some index set \(\mathcal T\). In this thesis, in the case that the index set is \(\mathcal T=\{0,1,2,...\}\) we say that the process \(\{X(n)\}_{n\in \mathcal T}\) is a discrete-time process, and we will typically use the dummy variable \(n\) for the `time' index. When the index set is \(\mathcal T=[0,\infty)\) we say \(\{X(t)\}_{t\in\mathcal T}\) is a continuous-time process, and we will typically use the dummy variable \(t\) for the `time' index. We may omit the index set and write \(\{X(t)\}\) in place of \(\{X(t)\}_{t\in\mathcal T}\) when it is not explicitly needed, or we may write \(\{X(t)\}_{t\geq t_0}\) to mean \(\{X(t)\}_{t\in[t_0,\infty)}\). The \emph{sample space} of \(X(t)\) is the set of possible values that \(X(t)\) can take at any time \(t\in\mathcal T\).

The \emph{initial distribution}, \(\mu\), of a stochastic process is the distribution of \(X(0)\). More generally, for a random variable, \(Z\), we write \(Z\sim \nu\) when \(Z\) has the distribution given by the probability measure \(\nu\). For the probability that \(X(t)\) lies in some set \(E\subset \mathcal S\) given \(X(0)\sim \mu\), we write \(\mathbb P(X(t)\in E \mid X(0)\sim \mu)\). When \(\mu\) assigns probability 1 to a single point, \(x\in\calS\), say, we write \(\mathbb P(X(t)\in E \mid X(0)=x).\)

A stochastic process is \emph{stationary} if
\begin{align}\label{eqn: stationary}
	&\nonumber\mathbb P(X(t_n)\in E_n,X(t_{n-1})\in E_{n-1},...,X(t_0)\in E_0) 
	\\&= \mathbb P(X(t_n+t)=i,X(t_{n-1}+t)\in E_{n-1},...,X(t_0+t)\in E_0),
\end{align}
for any \(t,t_0,...,t_n\in\mathcal T\) and any \(E_0,...,E_n\subseteq \mathcal S\), i.e.~the joint distribution~(\ref{eqn: stationary}) does not depend on \(t\).
A random variable, \(\tau\), is a \emph{stopping time} for the stochastic process \(\{X(t)\}_{t\in\mathcal T}\) if \(\tau\in\sigma(X(s), s\leq t),\) where \(\sigma(X(s), s\leq t)\) denotes the \(\sigma\)-algebra generated by \(\{X(s), s\leq t\}\).

We call \(\{X(n)\}_{n\in\{0,1,2,\dots\}}\) a \emph{discrete-time Markov chain} if \(\calS\) is countable and
\begin{align}\label{eqn: Markov property}
	\nonumber &\mathbb P(X(n+1)=j \mid X(n)=i,X(n-1)=i_{n-1},...,X(0)=i_0) 
	\\&= \mathbb P(X(n+1)=j \mid X(n)=i) 
\end{align}
for all \(n\in\{0,1,2,...\}\) and \(i,i_0,...,i_{n-1},j\in\calS,\) and refer to (\ref{eqn: Markov property}) as the Markov property. The process \(\{X(n)\}_{n\in\{0,1,2,...\}}\) is said to be time-homogeneous if 
\(\mathbb P(X(n+1)=j \mid X(n)=i)\) does not depend on \(n\). The \emph{strong Markov property} states that for each stopping time \(\tau\) of the Markov chain \(\{X(n)\}_{n\in\{0,1,2,...\}}\), on the event that \(\{\tau<\infty\}\), then 
\begin{align}\label{eqn: strong Markov property}
	\mathbb P(X(\tau+1)=j \mid \sigma(X(0),X(1),...,X(\tau))) = \mathbb P(X(\tau+1)=j \mid X(\tau)).
\end{align}

We call \(\{X(t)\}_{t\geq 0}\) a \emph{continuous-time Markov chain} if \(\calS\) is countable and for all \(t_{n+1} > t_n > t_{n-1} > ... > t_0 > 0\) and \(i,i_0,...,i_{n-1},j\in\calS,\)
\begin{align}\label{eqn: Markov property cmtc}
	\mathbb P(X(t_{n+1})=j \mid X(n)=i, X(n-1)=i_{n-1},...,X(0)=i_0) = \mathbb P(X(t_{n+1})=j \mid X(t_n)=i),
\end{align}
and refer to (\ref{eqn: Markov property cmtc}) as the \emph{Markov property}.
The process \(\{X(t)\}_{t\geq 0}\) is said to be time-homogeneous if 
\(\mathbb P(X(t+s)=j \mid X(s)=i)\) does not depend on \(s\). The \emph{strong Markov property} states that for each stopping time \(\tau\) of the Markov chain \(\{X(t)\}_{t\geq 0}\), on the event that \(\{\tau<\infty\}\), then 
\begin{align}\label{eqn: strong Markov property ctmc}
	\mathbb P(X(\tau + t)=j \mid \sigma(X(s),s\leq \tau)) = \mathbb P(X(\tau+t)=j \mid X(\tau)).
\end{align}

@@

For example, for a bounded linear operator \(\mathbb B\), define the operator exponential as \(e^{\mathbb Bt}:=\sum_{k=0}^\infty\cfrac{1}{k!}t^k\mathbb B^k\), then it can be shown that \(e^{\mathbb B(t+s)}=e^{\mathbb Bt}e^{\mathbb Bs}\), and from the definition \(e^{\mathbb B0}=I\), hence \(\{e^{\mathbb Bt}:t\geq 0\}\) is a semigroup. As another example, let \(\mathbb B=\cfrac{d}{dx}:C^\omega \to C^\omega(\mathbb R)\), where \(C^\omega(\mathbb R)\) is the class of analytic functions on the whole of \(\mathbb R\). Then \(\sum_{k=0}^\infty \cfrac{1}{k!}t^k\cfrac{\wrt^k}{\wrt x^k}f(x)=f(x-t)\), and the sum converges since \(f\in C^\omega(\mathbb R)\); the series representation is just the Taylor series of \(f\) about the point \(x\).

\subsubsection{Infinitesimal generators} The \emph{infinitesimal generator} (or just \emph{generator} for short) of a semigroup \(\{\mathbb T(t)\}\) is the linear operator defined by 
\[\mathbb Bf = \lim\limits_{t\to 0^+}\frac{1}{t}(\mathbb T(t)f-I),\]
and the domain, \(\mathcal D(\mathbb B)\), is the subspace of \(L\) for which this limit exists. Note that an essential part of the definition of the generator is its domain. For this reason it is common to denote the generator as the pair \((\mathbb B,\mathcal D(\mathbb B))\).

To determine the generator \(\mathbb B\) we can proceed by either differentiation of the semigroup, 
\[\mathbb Bf = \lim\limits_{t\to 0^+}\frac{1}{t}(\mathbb T(t)f-I),\]
or integration;
\[\int_{t=0}^\infty e^{-st}\mathbb T(t)\wrt t = (sI-\mathbb B)^{-1}=:R_s.\]
The operator \(R_s\) is known as the resolvent.

@@

\section{Fluid queues}\label{sec: fqs}
An unbounded fluid queue is a two-dimensional stochastic process which we denote by \(\{\ddot{\bs X}(t)\} = \{(\ddot X(t),\varphi(t))\}_{t\geq0}\) where \(\{\varphi(t)\}_{t\geq0}\) is known as the phase or driving process, and \(\{\ddot X(t)\}_{t\geq0}\) is known as the level process or buffer. The phase process \(\{\varphi(t)\}_{t\geq0}\), is an irreducible continuous-time Markov chain (CTMC) with finite state space, which we assume to be \(\mathcal S=\{1,2,\dots,N\}\), and infinitesimal generator \(\bs T= [T_{ij}]_{i,j\in\mathcal S}\). We assume that \(\bs T\) is \emph{conservative}. Associated with states \(i\in\mathcal S\) are real-valued \emph{rates} \(c_i\in\mathbb R\) which determine the rate at which \(\{X(t)\}\) moves.

Partition the state space \(\calS\) into \(\calS_+ = \{i\in\calS\mid c_i>0\}\), \(\calS_- = \{i\in\calS\mid c_i<0\}\), \(\calS_0 = \{i\in\calS\mid c_i=0\}\), \(\mathcal S_{-1}=\{i\in\calS\mid c_i\leq0\},\, \mathcal S_{K+1}=\{i\in\calS\mid c_i\geq0\}\)\footnote{The notation \(\mathcal S_{-1}\) and \(\mathcal S_{K+1}\) will make sense later when we introduce the partition into cells for the approximation schemes into \(K\) cells, plus a lower and upper boundary which we represent with indices \(-1\) and \(K+1\) respectively.}. We assume, without loss of generality, that the generator \(\bs T\) is partitioned into sub-matrices
\[\bs T = \left[\begin{array}{ccc}\bs T_{++} & \bs T_{+-} & \bs T_{+0} \\ \bs T_{-+} & \bs T_{--} & \bs T_{-0} \\ \bs T_{0+} & \bs T_{0-} & \bs T_{00}  \end{array}\right],\]
where \(\bs T_{mn} = [T_{ij}]_{i\in\mathcal S_m, j\in\mathcal S_n}\), \(m,n\in\{+,-,0\}\).\footnote{Let's clarify some notation. We use the notation \(\bs u = (u_h)_{h\in \mathcal H}\) to denote a row-vector, \(\bs u\), defined by its elements, \(u_h\), indexed by \(h\in\mathcal H\), where \(\mathcal H\) is some index set. Similarly, \(\bs u = (\bs u_h)_{h\in\mathcal H}\), is a row-vector defined by a collection of row-vectors \(\bs u_h\). The notation \(\bs u_m=(u_h)_{h\in\mathcal H_m}\) refers to the vector containing the subset of elements corresponding to \(\mathcal H_m\subseteq \mathcal H\). When the index set is empty, the resulting vector \(\bs u_m\) is a vector of dimension 0. In cases when there are two indices, we order the elements of the vector according to the first index, then the second; i.e.~\(\bs u = (u_{g}^h)_{g\in\mathcal G,h\in\mathcal H} = ((u_g^h)_{g\in\mathcal G})_{h\in\mathcal H}\). Here we use the convention that for a vector \(\bs u=(u)_{h\in \mathcal H}\) where the elements \(u\) do not depend on the index \(h\) and \(H\) is some index set, then we repeat \(u\) \(h\)-times; i.e.~\(\bs u = (u)_{h\in \mathcal H} =\underbrace{(u,\dots,u)}_{h-\mbox{times}}\). The notation \(\bs U = [u_{gh}]_{g\in\mathcal G, h\in\mathcal H}\) (square brackets) is used to denote a matrix defined by its elements, or sub-blocks, \(u_{gh}\).} Also define the diagonal matrices 
\begin{align*}
	\bs C &= \left[\begin{array}{ccc} \bs C_+ && \\ &\bs C_-& \\ && \bs 0\end{array}\right], && \bs C_+ = diag(c_i,i\in\calS_+), && \bs C_- = diag(|c_i|,i\in\calS_-),
\end{align*}
and 
\(
	\widehat{\bs C} = diag(c_i,i\in\calS),
\)
where \(diag(a_i,i\in\mathcal I)\) denotes a diagonal matrix with entries \(a_i\) down the diagonal.

@@

Often, boundary conditions are imposed. We denote a fluid queue bounded below at \(0\) and unbounded above by \(\{\dot{\bs X}(t)\} = \{(\dot X(t),\varphi(t))\}_{t\geq0}\), and a fluid queue bounded below at \(0\) and above by \(b<\infty\) by \(\{{\bs X}(t)\} = \{( X(t),\varphi(t))\}_{t\geq0}\). Here, we consider a mixture of \emph{regulated} and \emph{reflecting} boundary conditions. Upon hitting a boundary we suppose that, with probability \(p_{ij},\,i,j\in\mathcal S\), the phase process instantaneously transitions from phase \(i\) to phase \(j\) (note that we might have \(i=j\) i.e.~no transition) and if \(sign(c_i)=sign(c_j)\) or \(sign(c_j)=0\) then the process is absorbed in the boundary, otherwise it is reflected. At a lower boundary, if \(j\in\calS_0\cup\calS_-\), then \(\cfrac{\wrt}{\wrt t} X(t) = 0\), and the phase process continues to evolve according to the sub-generator 
\[\left[\begin{array}{cc} \bs T_{--} & \bs T_{-0} \\ \bs T_{0-} & \bs T_{00}  \end{array}\right],\]
until such a time that \(\{\varphi(t)\}\) transitions to a phase \(k\in\calS_+\), at which time \(\{X(t)\}\) leaves the boundary. Similarly, at an upper boundary if \(j\in\calS_0\cup\calS_+\), then \(\cfrac{\wrt}{\wrt t} X(t) = 0\) and the phase process continues to evolve according to the sub-generator 
\[\left[\begin{array}{cc} \bs T_{++} & \bs T_{+0} \\ \bs T_{0+} & \bs T_{00}  \end{array}\right],\]
until such a time that \(\{\varphi(t)\}\) transitions to a phase \(k\in\calS_-\) at which time \(\{X(t)\}\) leaves the boundary. Without loss of generality, we assume the lower and upper boundaries (when present) are at \(x=0\) and \(x=b>0\), respectively.

@@

Let \(\bs f(x,t) = (f_i(x,t))_{i\in\calS}\) be a row-vector function where \(f_i(x,t)\) is the density of \(\mathbb P(X(t)\leq x, \varphi(t) = i\mid \bs X(0)\sim \bs \mu )\), assuming it exists. When a differentiable density exists, the system of partial differential equation which describes the evolution of the densities \(\bs f(x,t)\) is 
\begin{align}
	\cfrac{\partial}{\partial t} \bs f(x,t) = \bs f(x,t)\bs T - \cfrac{\partial}{\partial x}\bs f(x,t)\widehat{\bs C},\label{eqn: pde}
\end{align}
on the interior \(x\in(0,b)\), with appropriate boundary conditions (see Section~\ref{subsec: boundary DG}). The initial condition is the initial distribution of the fluid queue which, when it has a density, we write as \(f_i(x,0)\). Often a differentiable density function does not exist and therefore the partial differential equation~(\ref{eqn: pde}) is not well-defined. For example, for a fluid queue with no upper boundary, if the initial distribution of the fluid queue is a point mass at any point \(x_0\geq 0\) and in phase \(i\in\calS_+\cup\calS_0\), then a density function \(f_i(x,t)\) will not exist for any finite \(t\). Specifically, a point mass will persist along the ray \(x_0+c_it\), \(t\geq 0\). In such situations, it is the \emph{weak solution} to (\ref{eqn: pde}) that we seek. A weak solution satisfies
\begin{align}
	-\int_{x=0}^b\int_{t=0}^\infty \bs f(x,t)\cfrac{\partial}{\partial t} \bs \psi(x,t)\wrt t \wrt x= &\int_{x=0}^b\int_{t=0}^\infty\bs f(x,t)\bs T\bs \psi(x,t)\wrt t \wrt x \nonumber 
	\\&{} + \int_{x=0}^b \int_{t=0}^\infty\bs f(x,t)\widehat{\bs C} \cfrac{\partial}{\partial x} \bs \psi(x,t)\wrt t \wrt x,\label{eqn: weak pde}
\end{align}
for every row-vector of test functions, \(\bs \psi(x,t) = ( \psi_i(x,t))_{i\in\calS}\), which are smooth, have compact support and \(\bs \psi(x,0)=\bs \psi(0,t) = \bs \psi(b,t)=\bs 0\).

\subsection{Transient analysis of fluid queues}\label{sec: transient ffq intro}
The transient analysis of fluid queues has been relatively well studied (see, for example, \cite{ar2004,bean2005,dasilva2005,bean2009} which are matrix-analytics-methods-based, and also \cite{rs2003} which is differential-equations-based and treats a slightly more complex model than the ones considered here, but is none-the-less relevant). Given the interest in discontinuous and point-mass initial conditions, transient analysis of fluid queues rarely relies on solving governing differential equations numerically, although it is quite possible to do so\footnote{We have already alluded to some difficulties with differential-equation-based approaches to problems with discontinuous solutions such as ill-defined PDEs, the need for weak solutions, and oscillatory approximations with possibly infeasible approximate solutions, for example negative probability approximations.}. Instead, expressions for transient distributions of fluid queues are derived in terms of Laplace transforms with respect to time, and/or moments of the transient distributions are derived \citep{ar2004,bean2005,bean2009}. Moreover, these techniques often lead to quantities which have direct probabilistic interpretations which further aids in their analysis and also applications to other problems \citep{ar2003,dasilva2005,bnp2018} (we also leverage these stochastic interpretations in Chapter~\ref{sec: conv} to derive expressions for certain the Laplace transforms of the fluid queue). In some context, the actual transient distributions are not required, and Laplace transforms or moments are all that are required. In other cases, we can invert the Laplace transforms using known methods (\cite{aw2006}, or in the case of the discontinuities we might prefer \cite{hhat2020}, which uses the same concentrated matrix exponential distribution that we do).

@@

\subsection{Stationary Distribution}

>>>>

Following \cite{rossBook}, a stochastic process is a sequence, \(\{X(t)\}_{t\in\mathcal T}\), of random variables (or random vectors) indexed by some index set \(\mathcal T\). In this thesis, in the case that the index set is \(\mathcal T=\{0,1,2,...\}\) we say that the process \(\{X(n)\}_{n\in \mathcal T}\) is a discrete-time process, and we will typically use the dummy variable \(n\) for the `time' index. When the index set is \(\mathcal T=[0,\infty)\) we say \(\{X(t)\}_{t\in\mathcal T}\) is a continuous-time process, and we will typically use the dummy variable \(t\) for the `time' index. We may omit the index set and write \(\{X(t)\}\) in place of \(\{X(t)\}_{t\in\mathcal T}\) when it is not explicitly needed, or we may write \(\{X(t)\}_{t\geq t_0}\) to mean \(\{X(t)\}_{t\in[t_0,\infty)}\). The \emph{state space} of \(X(t)\) is the set of possible values that \(X(t)\) can take at any time \(t\in\mathcal T\).

The \emph{initial distribution}, \(\mu\), of a stochastic process is the distribution of \(X(0)\). More generally, for a random variable, \(Z\), we write \(Z\sim \nu\) when \(Z\) has the distribution given by the probability measure \(\nu\). For the probability that \(X(t)\) lies in some measurable set \(E\subset \mathcal S\) given \(X(0)\sim \mu\), we write \(\mathbb P(X(t)\in E \mid X(0)\sim \mu)\). When \(\mu\) assigns probability 1 to a single point, \(x\in\calS\), say, we write \(\mathbb P(X(t)\in E \mid X(0)=x).\)

A stochastic process is \emph{stationary} if \(X(t_n),X(t_{n-1}),...,X(t_0)\) and \(X(t_n+t),X(t_{n-1}+t),...,X(t_0+t)\) have the same distribution for any \(t\). That is, 
\begin{align}\label{eqn: stationary}
	&\nonumber\mathbb P(X(t_n)\in E_n,X(t_{n-1})\in E_{n-1},...,X(t_0)\in E_0) 
	\\&= \mathbb P(X(t_n+t)\in E_n,X(t_{n-1}+t)\in E_{n-1},...,X(t_0+t)\in E_0),
\end{align}
for any \(t,t_0,...,t_n\in\mathcal T\) with \(t_0+t,...,t_n+t\in\mathcal T\) and any \(E_0,...,E_n\subseteq \mathcal S\). A random variable, \(\tau\), is a \emph{stopping time} for the stochastic process \(\{X(t)\}_{t\in\mathcal T}\) if \(\tau\in\sigma(X(s), s\leq t),\) where \(\sigma(X(s), s\leq t)\) denotes the \(\sigma\)-algebra generated by \(\{X(s), s\leq t\}\).

As in \cite[Chapter~1.2]{MEinAP}, we call \(\{X(n)\}_{n\in\{0,1,2,\dots\}}\) a \emph{discrete-time Markov chain} if \(\calS\) is countable and
\begin{align}\label{eqn: Markov property}
	\nonumber &\mathbb P(X(n+1)=j \mid X(n)=i,X(n-1)=i_{n-1},...,X(0)=i_0) 
	\\&= \mathbb P(X(n+1)=j \mid X(n)=i) 
\end{align}
for all \(n\in\{0,1,2,...\}\) and \(i,i_0,...,i_{n-1},j\in\calS,\) and refer to (\ref{eqn: Markov property}) as the Markov property. The process \(\{X(n)\}_{n\in\{0,1,2,...\}}\) is said to be time-homogeneous if 
\(\mathbb P(X(n+1)=j \mid X(n)=i)\) does not depend on \(n\). The probabilities \(\mathbb P(X(n+1)=j \mid X(n)=i)=:p_{ij}\) are the \emph{transition probabilities}, and \(\bs P=[p_{ij}]_{i,j\in\calS}\) is the \emph{transition matrix}, of the Markov chain. The \emph{strong Markov property} states that for each stopping time \(\tau\) of the Markov chain \(\{X(n)\}_{n\in\{0,1,2,...\}}\), on the event that \(\{\tau<\infty\}\), then 
\begin{align}\label{eqn: strong Markov property}
	\mathbb P(X(\tau+1)=j \mid \sigma(X(0),X(1),...,X(\tau))) = \mathbb P(X(\tau+1)=j \mid X(\tau)).
\end{align}

Also, as in \cite[Section~1.3]{MEinAP}, we call \(\{X(t)\}_{t\geq 0}\) a \emph{continuous-time Markov chain} (CTMC) if \(\calS\) is countable and for all \(t_{n+1} > t_n > t_{n-1} > ... > t_0 > 0\) and \(i,i_0,...,i_{n-1},j\in\calS,\)
\begin{align}\label{eqn: Markov property cmtc}
	\mathbb P(X(t_{n+1})=j \mid X(n)=i, X(n-1)=i_{n-1},...,X(0)=i_0) = \mathbb P(X(t_{n+1})=j \mid X(t_n)=i),
\end{align}
and refer to (\ref{eqn: Markov property cmtc}) as the \emph{Markov property}. The process \(\{X(t)\}_{t\geq 0}\) is said to be time-homogeneous if 
\(\mathbb P(X(t+s)=j \mid X(s)=i)\) does not depend on \(s\). For a time-homogeneous CTMC, the \emph{transition function} is the matrix function 
\[\bs P(t) = [\mathbb P(X(t)=j \mid X(t)=i)]_{i,j\in\calS}.\]
The Chapman-Kolmogorov equations state that 
\begin{align} 
	&\nonumber \bs P(t+s)=\bs P(t)\bs P(s).
\end{align}
The \emph{infinitesimal generator} matrix of a CTMC is 
\begin{align} 
	&\nonumber \bs T = [T_{ij}]_{i,j\in\calS} = \cfrac{\wrt}{\wrt t}P(t),
\end{align}
and the elements \(T_{ij}\) are the \emph{transition rates}. A CTMC is said to be \emph{conservative} if \(\bs P(t)\bs 1=\bs 1\) for all \(t\geq 0\) where \(\bs 1\) is a column vector of ones.

The \emph{strong Markov property} states that for each stopping time \(\tau\) of the Markov chain \(\{X(t)\}_{t\geq 0}\), on the event that \(\{\tau<\infty\}\), then 
\begin{align}\label{eqn: strong Markov property ctmc}
	\mathbb P(X(\tau + t)=j \mid \sigma(X(s),s\leq \tau)) = \mathbb P(X(\tau+t)=j \mid X(\tau)).
\end{align}

@@

For example, for a bounded linear operator \(\mathbb B:\mathcal D(\mathbb B)\to \mathcal D(\mathbb B)\), define the operator exponential as \(e^{\mathbb Bt}:=\sum_{k=0}^\infty\cfrac{1}{k!}t^k\mathbb B^k\) where, for \(f\in \mathcal D(\mathbb B)\), \(\mathbb B^kf\) is defined recursively by \(\mathbb B^k f = \mathbb B^{k-1} (\mathbb B f)\). It can be shown that \(e^{\mathbb B(t+s)}=e^{\mathbb Bt}e^{\mathbb Bs}\), and from the definition \(e^{\mathbb B0}=I\), hence \(\{e^{\mathbb Bt}:t\geq 0\}\) is a semigroup. As another example, let \(\mathbb B=\cfrac{d}{dx}:C^\omega(\mathbb R) \to C^\omega(\mathbb R)\), where \(C^\omega(\mathbb R)\) is the class of analytic functions on the whole of \(\mathbb R\). Then \(\mathbb T(t)f(x)=\sum_{k=0}^\infty \cfrac{1}{k!}t^k\cfrac{\wrt^k}{\wrt x^k}f(x)=f(x-t)\), and the sum converges since \(f\in C^\omega(\mathbb R)\); the series representation is the Taylor series of \(f\) about the point \(x\) and the operator \(\mathbb T(t)f(x)=f(x-t)\) is the shift operator. The semigroup property follows since 
\[\mathbb T(t)\mathbb T(s)f(x)=\mathbb T(t)f(x-s)=f(x-s-t)=\mathbb T(t+s)f(x).\]

\subsubsection{Infinitesimal generators} The \emph{infinitesimal generator} (or just \emph{generator} for short) of a semigroup \(\{\mathbb T(t)\}\) is the linear operator defined by 
\[\mathbb Bf = \lim\limits_{t\to 0^+}\frac{1}{t}(\mathbb T(t)f-f),\]
and the domain, \(\mathcal D(\mathbb B)\), is the subspace of \(L\) for which this limit exists. Note that an essential part of the definition of the generator is its domain. For this reason it is common to denote the generator as the pair \((\mathbb B,\mathcal D(\mathbb B))\).

To determine the generator \(\mathbb B\) we can proceed by either differentiation of the semigroup, 
\[\mathbb Bf = \lim\limits_{t\to 0^+}\frac{1}{t}(\mathbb T(t)f-f),\]
or integration;
\[\int_{t=0}^\infty e^{-st}\mathbb T(t)\wrt t = (sI-\mathbb B)^{-1}=:R_s.\]
The operator \(R_s\) is known as the resolvent.

@@

\section{Fluid queues}\label{sec: fqs}
An unbounded fluid queue is a two-dimensional stochastic process which we denote by \(\{\ddot{\bs X}(t)\} = \{(\ddot X(t),\varphi(t))\}_{t\geq0}\) where \(\{\varphi(t)\}_{t\geq0}\) is known as the phase or driving process, and \(\{\ddot X(t)\}_{t\geq0}\) is known as the level process or buffer. The phase process \(\{\varphi(t)\}_{t\geq0}\), is an irreducible continuous-time Markov chain (CTMC) with finite state space, which we assume to be \(\mathcal S=\{1,2,\dots,N\}\), and infinitesimal generator \(\bs T= [T_{ij}]_{i,j\in\mathcal S}\). We assume that \(\bs T\) is {conservative} and time-homogeneous. Associated with states \(i\in\mathcal S\) are real-valued \emph{rates} \(c_i\in\mathbb R\) which determine the rate at which \(\{X(t)\}\) moves.

Partition the state space \(\calS\) into \(\calS_+ = \{i\in\calS\mid c_i>0\}\), \(\calS_- = \{i\in\calS\mid c_i<0\}\), \(\calS_0 = \{i\in\calS\mid c_i=0\}\), \(\mathcal S_{-1}=\{i\in\calS\mid c_i\leq0\},\, \mathcal S_{K+1}=\{i\in\calS\mid c_i\geq0\}\)\footnote{The notation \(\mathcal S_{-1}\) and \(\mathcal S_{K+1}\) will make sense later when we introduce the partition into cells for the approximation schemes into \(K\) cells, plus a lower and upper boundary which we represent with indices \(-1\) and \(K+1\) respectively.}. We assume, without loss of generality, that the generator \(\bs T\) is partitioned into sub-matrices
\[\bs T = \left[\begin{array}{ccc}\bs T_{++} & \bs T_{+-} & \bs T_{+0} \\ \bs T_{-+} & \bs T_{--} & \bs T_{-0} \\ \bs T_{0+} & \bs T_{0-} & \bs T_{00}  \end{array}\right],\]
where \(\bs T_{mn} = [T_{ij}]_{i\in\mathcal S_m, j\in\mathcal S_n}\), \(m,n\in\{+,-,0\}\).

{Let's clarify some notation. We use the notation \(\bs u = (u_h)_{h\in \mathcal H}\) to denote a row-vector, \(\bs u\), defined by its elements, \(u_h\), indexed by \(h\in\mathcal H\), where \(\mathcal H\) is some index set. Similarly, \(\bs u = (\bs u_h)_{h\in\mathcal H}\), is a row-vector defined by a collection of row-vectors \(\bs u_h\). The notation \(\bs u_m=(u_h)_{h\in\mathcal H_m}\) refers to the vector containing the subset of elements corresponding to \(\mathcal H_m\subseteq \mathcal H\). When the index set is empty, the resulting vector \(\bs u_m\) is a vector of dimension 0. In cases when there are two indices, we order the elements of the vector according to the first index, then the second; i.e.~\(\bs u = (u_{g}^h)_{g\in\mathcal G,h\in\mathcal H} = ((u_g^h)_{g\in\mathcal G})_{h\in\mathcal H}\). Here we use the convention that for a vector \(\bs u=(u)_{h\in \mathcal H}\) where the elements \(u\) do not depend on the index \(h\) and \(H\) is some index set, then we repeat \(u\) \(h\)-times; i.e.~\(\bs u = (u)_{h\in \mathcal H} =\underbrace{(u,\dots,u)}_{h-\mbox{times}}\). The notation \(\bs U = [u_{gh}]_{g\in\mathcal G, h\in\mathcal H}\) (square brackets) is used to denote a matrix defined by its elements, or sub-blocks, \(u_{gh}\).}

Also define the diagonal matrices 
\begin{align*}
	\bs C &= \left[\begin{array}{ccc} \bs C_+ && \\ &\bs C_-& \\ && \bs 0\end{array}\right], && \bs C_+ = diag(c_i,i\in\calS_+), && \bs C_- = diag(|c_i|,i\in\calS_-),
\end{align*}
and 
\(
	\widehat{\bs C} = diag(c_i,i\in\calS),
\)
where \(diag(a_i,i\in\mathcal I)\) denotes a diagonal matrix with entries \(a_i\) down the diagonal.

@@

Often, boundary conditions are imposed. We denote a fluid queue bounded below at \(0\) and unbounded above by \(\{\dot{\bs X}(t)\} = \{(\dot X(t),\varphi(t))\}_{t\geq0}\), and a fluid queue bounded below at \(0\) and above at \(b<\infty\) by \(\{{\bs X}(t)\} = \{( X(t),\varphi(t))\}_{t\geq0}\). Here, we consider a mixture of \emph{regulated} and \emph{reflecting} boundary conditions. Upon hitting a boundary we suppose that, with probability \(p_{ij},\,i,j\in\mathcal S\), the phase process instantaneously transitions from phase \(i\) to phase \(j\) (note that we might have \(i=j\) i.e.~no transition) and if \(sign(c_i)=sign(c_j)\) or \(sign(c_j)=0\) then the process is absorbed in the boundary, otherwise it is reflected. At a lower boundary, if \(j\in\calS_0\cup\calS_-\), then \(\cfrac{\wrt}{\wrt t} X(t) = 0\), and the phase process continues to evolve according to the sub-generator 
\[\left[\begin{array}{cc} \bs T_{--} & \bs T_{-0} \\ \bs T_{0-} & \bs T_{00}  \end{array}\right],\]
until such a time that \(\{\varphi(t)\}\) transitions to a phase \(k\in\calS_+\), at which time \(\{X(t)\}\) leaves the boundary. Similarly, at an upper boundary if \(j\in\calS_0\cup\calS_+\), then \(\cfrac{\wrt}{\wrt t} X(t) = 0\) and the phase process continues to evolve according to the sub-generator 
\[\left[\begin{array}{cc} \bs T_{++} & \bs T_{+0} \\ \bs T_{0+} & \bs T_{00}  \end{array}\right],\]
until such a time that \(\{\varphi(t)\}\) transitions to a phase \(k\in\calS_-\) at which time \(\{X(t)\}\) leaves the boundary. It is without loss of generality that we assume the lower and upper boundaries (when present) are at \(x=0\) and \(x=b>0\), respectively.

@@

Let \(\bs f(x,t) = (f_i(x,t))_{i\in\calS}\) be a row-vector function where \(f_i(x,t)\) is the density of \(\mathbb P(X(t)\leq x, \varphi(t) = i\mid \bs X(0)\sim \bs \mu )\), assuming it exists. When a differentiable density exists, the system of partial differential equations which describes the evolution of the densities \(\bs f(x,t)\) is 
\begin{align}
	\cfrac{\partial}{\partial t} \bs f(x,t) = \bs f(x,t)\bs T - \cfrac{\partial}{\partial x}\bs f(x,t)\widehat{\bs C},\label{eqn: pde}
\end{align}
on the interior \(x\in(0,b)\), with appropriate boundary conditions (see Section~\ref{subsec: boundary DG}). The initial condition is the initial distribution of the fluid queue which, when it has a density, we write as \(f_i(x,0)\). Often a differentiable density function does not exist and therefore the partial differential equation~(\ref{eqn: pde}) is not well-defined. For example, for a fluid queue with no upper boundary, if the initial distribution of the fluid queue is a point mass at any point \(x_0\geq 0\) and in phase \(i\in\calS_+\cup\calS_0\), then a density function \(f_i(x,t)\) will not exist for any finite \(t\). Specifically, a point mass will persist along the ray \(x_0+c_it\), \(t\geq 0\). In such situations, it is the \emph{weak solution} to (\ref{eqn: pde}) that we seek. A weak solution (ignoring boundary conditions) satisfies
\begin{align}
	-\int_{x=0}^b\int_{t=0}^\infty \bs f(x,t)\cfrac{\partial}{\partial t} \bs \psi(x,t)\wrt t \wrt x= &\int_{x=0}^b\int_{t=0}^\infty\bs f(x,t)\bs T\bs \psi(x,t)\wrt t \wrt x \nonumber 
	\\&{} + \int_{x=0}^b \int_{t=0}^\infty\bs f(x,t)\widehat{\bs C} \cfrac{\partial}{\partial x} \bs \psi(x,t)\wrt t \wrt x,\label{eqn: weak pde}
\end{align}
for every row-vector of test functions, \(\bs \psi(x,t) = ( \psi_i(x,t))_{i\in\calS}\), which are smooth, have compact support and \(\bs \psi(x,0)=\bs \psi(0,t) = \bs \psi(b,t)=\bs 0\) \cite[Chapter~10]{borthwickBook}.

\subsection{Transient analysis of fluid queues}\label{sec: transient ffq intro}
The transient analysis of fluid queues has been relatively well studied (see, for example, \cite{ar2004,bean2005,dasilva2005,bean2009} which are matrix-analytics-methods-based, and also \cite{rs2003} which is differential-equations-based and treats a slightly more complex model than the ones considered here, but is none-the-less relevant). Given the interest in discontinuous and point-mass initial conditions, transient analysis of fluid queues rarely relies on solving governing differential equations numerically, although it is quite possible to do so\footnote{We have already alluded to some difficulties with differential-equation-based approaches to problems with discontinuous solutions such as ill-defined PDEs, the need for weak solutions, and oscillatory approximations with possibly infeasible approximate solutions, for example negative probability approximations.}. Instead, expressions for transient distributions of fluid queues are derived in terms of Laplace transforms with respect to time, and/or moments of the transient distributions are derived \citep{ar2004,bean2005,bean2009}. Moreover, these techniques often lead to quantities which have direct probabilistic interpretations which further aids in their analysis and also applications to other problems \citep{ar2003,dasilva2005,bnp2018} (we also leverage these stochastic interpretations in Chapter~\ref{sec: conv} to derive expressions for certain the Laplace transforms of the fluid queue). In some contexts, the actual transient distributions are not required, and Laplace transforms or moments are all that are required. In other cases, we can invert the Laplace transforms using known methods (\cite{aw2006}, or in the case of the discontinuities we might prefer \cite{hhat2020}, which uses the same concentrated matrix exponential distribution that we do).

@@

\subsection{Limiting Distribution}
